---
title: "Covid_19_deaths_JH_data"
author: "JG"
date: "2023-08-22"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---
---
Title: Covid 19
Author: JG
Date: "2023-08-19"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

# Setup

### install.packages("tidyverse"), 
install.packages("readr"), install.packages("knitr"), 
install.packages ("stringr"), install.packages("dplyr") 

```{r, setup, warning=FALSE}
library(tidyverse) 
library(lubridate) 
library(stringr) 
library(readr) 
library(selectr)
```
# Import Data
### note:website error prohibited importing directly from site, save to local directory
```{r, import data, results=FALSE, warning=FALSE}
global_cases_url<-"C:\\Temp\\time_series_covid19_confirmed_global.csv"
global_deaths_url<-"C:\\Temp\\time_series_covid19_deaths_global.csv"
US_cases_url<-"C:\\Temp\\time_series_covid19_confirmed_US.csv"
US_deaths_url<-"C:\\Temp\\time_series_covid19_deaths_US.csv"
global_cases<-read_csv(global_cases_url)
global_deaths<-read_csv(global_deaths_url)
US_cases<-read_csv(US_cases_url)
US_deaths<-read_csv(US_deaths_url)
```


# Tidy data

```{r, tidy_global_data, results=FALSE, warning=FALSE}
global_cases<-global_cases %>% 
  pivot_longer(cols=-c('Province/State', 
                       'Country/Region', Lat, Long), 
              names_to="date",
              values_to="cases") %>% 
  select(-c(Lat,Long))
 
global_deaths<-global_deaths %>%
  pivot_longer(cols=-c('Province/State',
                       'Country/Region', Lat, Long),
               names_to="date",
               values_to="deaths") %>%
  select(-c(Lat,Long))
  
global<-global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region='Country/Region',
         Province_State='Province/State') %>%
  mutate(date=mdy(date))

US_cases<-US_cases %>%
  pivot_longer(cols=-(UID:Combined_Key),
               names_to="date",
               values_to="cases") %>%
    select(Admin2:cases) %>%
  mutate(date=mdy(date))%>%
    select(-c(Lat,Long_))
    
US_deaths<-US_deaths %>%
  pivot_longer(cols =-(UID:Population),
               names_to="date",
               values_to="deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(date))%>%
  select(-c(Lat,Long_))
```

```{r, US cases, results=FALSE, warning=FALSE}
US<-US_cases %>%
  full_join(US_deaths)
```

```{r, global, results=FALSE, warning=FALSE}
global<-global %>%
    unite("Combined_Key",
         c(Province_State, Country_Region),
         sep = ", ",
         na.rm=TRUE,
         remove=FALSE)
```

# Import population data

```{r, add_population_data, results=FALSE, warning=FALSE}
#add population data
uid_lookup_url<-"C:\\Temp\\UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <-global %>%
  left_join(uid, by=c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, 
         cases, deaths, Population, 
         Combined_Key)
```

# Visualize

```{r, ##visualize, results=FALSE, warning=FALSE}
##Visualize
US_by_state<-US %>%
    group_by(Province_State, Country_Region, date) %>%
    summarize(cases=sum(cases), deaths =sum(deaths),
              Population=sum(Population)) %>%
  mutate(deaths_per_mill=deaths *1000000/Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
US_totals<-US_by_state %>%
    group_by(Country_Region, date) %>%
    summarize(cases =sum(cases), deaths=sum(deaths),
              Population =sum(Population)) %>%
    mutate(deaths_per_mill=deaths *1000000/Population) %>%
    select(Country_Region, date,
           cases, deaths, deaths_per_mill, Population) %>%
    ungroup()
```
```{r, plot(US_totals), warning=FALSE}
US_totals %>%
    filter(cases>0) %>%
    ggplot(aes(x=date,y=cases))+
    geom_line(aes(color = "cases")) +
    geom_point(aes(color = "cases")) +
    geom_line(aes(y=deaths,color="deaths")) +
    geom_point(aes(y=deaths, color = "deaths")) +
    scale_y_log10() +
    theme(legend.position="bottom",
          axis.text.x=element_text(angle=90)) +
    labs(title="Covid-19 in US", y=NULL)
```
```{r, plot(US_by_state), warning=FALSE}
state<-"New York" 
US_by_state %>% 
  filter(Province_State==state) %>%  
  filter(cases>0) %>% 
  ggplot(aes(x=date,y=cases))+ 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) +
  geom_line(aes(y=deaths,color="deaths")) + 
  geom_point(aes(y=deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position ="bottom", 
        axis.text.x=element_text(angle=90)) + 
  labs(title = str_c("COVID19 in", state), y=NULL) 
US_by_state <-US_by_state %>% 
  mutate(new_cases=cases-lag(cases), 
         new_deaths=deaths-lag(deaths)) 
US_totals<-US_totals %>% 
  mutate(new_cases =cases-lag(cases), 
         new_deaths=deaths - lag (deaths)) 
tail(US_totals %>% 
       select(new_cases, new_deaths,everything())) 
US_totals %>% 
  ggplot(aes(x=date, y=new_cases)) + 
  geom_line(aes(color="new cases")) + 
  geom_point(aes(color="new_cases")) + 
  geom_line(aes(y=new_deaths, color = "new_deaths")) +
  geom_point(aes(y=new_deaths, color ="new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position ="bottom", 
        axis.text.x=element_text(angle=90)) + 
  labs(title="COVID19 in US", y=NULL) 
US_states_totals<- US_by_state %>% 
  group_by(Province_State) %>% 
  summarize(deaths=max(deaths), 
            cases=max(cases), population=max(Population),
            cases_per_thou=1000*cases/population,
            deaths_per_thou=1000*deaths/population) %>%
  filter(cases>0, population>0) 

US_states_totals %>% 
  slice_min(deaths_per_thou, n=10) %>% 
  select(deaths_per_thou, cases_per_thou, everything())
```

# Model
```{r, model, results=FALSE, warning=FALSE}
mod<-lm(deaths_per_thou~cases_per_thou, data=US_states_totals) 
summary(mod) 
US_states_totals %>% slice_min(cases_per_thou) 
US_states_totals %>% slice_max(cases_per_thou) 
x_grid<-seq(25,150)
new_df<-tibble(cases_per_thou =x_grid) 
US_tot_w_pred<-US_states_totals %>% mutate(pred=predict(mod))
US_tot_w_pred

US_tot_w_pred %>% 
  ggplot()+
geom_point(aes(x=cases_per_thou,y=deaths_per_thou), color="blue") + 
  geom_point(aes(x=cases_per_thou, y=pred), color="red")
```